OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2021-09-07 paper/paper.Rmd
Incorporate the rest of @pschloss' suggestions
    - Fixed incorrect assertion about Opti[Fit/Clust] requiring/enforcing the distance threshold.
    - Edited to use "cluster", "assign", "map", & "fit" in a consistent way. Mainly just use cluster.
    - Made the effects of greedy clustering vs opticlust more obvious without explicitly using the words "radius" or "diameter", since those haven't been used elsewhere.

<!-- TODO
- Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
1. The threshold with closed reference is a radius and with de novo its a diameter
2. The gg reference was defined for full length sequences, within a subregion there are duplicate and nearly duplicate sequences
3. Because VSEARCH/USEARCH is a greedy heuristic, its looking for a reference that is close. Although the ordering of the database is not relevant (it was with USEARCH) it gets the wrong number of its (see Figure 5)
-->

<!-- TODO
- "require" and "enforce" for opti -- wrong word choice.
- be consistent with cluster, assign, fit, and map.
- can we replace "robust" with something better / more specific?
-->

@@

Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
important step in characterizing the composition of microbial communities across
large datasets.
OptiClust, a _de novo_ OTU clustering method, has been
shown to produce higher quality OTU assignments than other methods and at
comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent
reference clustering methods is that OTU assignments from _de novo_ methods may
change when new sequences are added to a dataset.
However, in some cases one may wish to incorporate new samples into a previously
clustered dataset without performing clustering again on all sequences, such as
when comparing across datasets or deploying machine learning models where OTUs
are features.
Existing reference-based clustering methods produce consistent OTUs, but they
only consider the similarity of each query sequence to a single reference
sequence in an OTU, thus resulting in some dissimilar sequences being assigned
to the same OTU.
To provide an efficient and robust method to fit amplicon sequence data to
existing OTUs, we developed the OptiFit algorithm.
Like OptiFit, OptiClust considers the similarity of all pairs of reference and
query sequences in an OTU to produce OTUs of the best possible quality.
We tested OptiFit using four microbiome datasets with two different strategies:
by fitting to an external reference database or by splitting the dataset into a 
reference and query set and fitting the query sequences to the reference set
after clustering it using OptiClust.
The result is an improved implementation of closed and open-reference clustering.
OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
using the split dataset strategy, although the OTU quality and processing speed
depends on the database chosen when using the external database strategy.
OptiFit provides a suitable option for users who require consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering algorithm which uses the distance score
between all pairs of sequences in the dataset to cluster them into OTUs by
maximizing the Matthews Correlation Coefficient (MCC)
[@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_
methods such as the greedy clustering algorithms implemented in USEARCH and
VSEARCH, which only consider the distance between the query sequence and a
representative centroid sequence in the OTU [@edgar_search_2010;
@rognes_vsearch_2016].
In methods employing greedy clustering algorithms, some pairs of sequences in
the same OTU may have a greater distance than the specified threshold since only
the distance between each sequence and the centroid sequence is considered while
clustering.
In contrast, the OptiClust algorithm enforces that all pairs of sequences must
be within the distance threshold. <!-- i.e. radius vs. diameter -->
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Additionally, the greedy clustering algorithms are sensitive to the order of the
input sequences: different OTU assignments are produced when the same sequences
are randomly shuffled [@westcott_novo_2015; @he_stability_2015].
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow for
very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
Query sequences are then assigned to OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton. 
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
The MCC uses all values from a confusion matrix and ranges from zero to one,
with a score of one occurring when all sequence pairs are true positives and
true negatives, and a score of zero when all pairs are false positives and false
negatives.
Sequence pairs that are similar to each other (i.e. within the distance
threshold) are counted as true positives if they are assigned to the same OTU,
and false negatives if they are not in the the same OTU.
Sequence pairs that are not similar to each other are true negatives if they are
not assigned to the same OTU, and false positives if they are not in the same
OTU.
OptiClust iterations continue until the MCC stabilizes or until a maximum number
of iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.

@@

OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03) 
(Figure \ref{fig:algorithm}).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
All query and reference sequence pairs are considered when calculating the MCC.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
reference OTUs are discarded, and the results only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.

@@

While _de novo_ clustering produces the highest quality OTUs, researchers may
prefer to perform reference clustering to a public database because
reference-based methods produce consistent OTUs and are generally faster than
_de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% using the V4 region of each
sequence (see Figure \ref{fig:workflow}).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
respectively.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
was `r closed_fit_rdp_mcc` when fitting to RDP 
(see Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

@@

Since closed reference clustering does not cluster query sequences that could
not be assigned to reference OTUs, an additional measure of clustering
performance to consider is the fraction of query sequences that were able to be
assigned.
On average, more sequences were assigned with Greengenes as the reference 
(`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or with the RDP 
(`r frac_fit_rdp`%).
This mirrored the result reported above that Greengenes produced better OTUs in
terms of MCC score than either SILVA or RDP.
Note that _de novo_ and open reference clustering methods always assign 100% of
sequences to OTUs.
The database chosen affects the final OTU assignments considerably in terms of
both MCC score and fraction of query sequences that could be fit to the
reference OTUs.

@@

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH in terms of MCC, but performed 
`r sec_opticlust_vs_vsearch`% slower than VSEARCH.
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH, but VSEARCH was able to map 
`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
reference database.
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of sequences in an OTU. <!-- VSEARCH "diameter" vs OptiFit "radius"
-->
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively).
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average respectively.
Thus, the more stringent OTU definition employed by OptiFit, which requires the
query to be similar to all other sequences in the OTU rather than to one
sequence, resulted in fewer sequences being fit to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTU quality may be poor when the reference database consists of sequences that
are too unrelated to the samples of interest, such as when samples contain novel
populations.
While _de novo_ clustering overcomes the quality limitations of reference
clustering to databases, OTU assignments are not consistent when new sequences
are added.
Researchers may wish to fit new sequences to existing OTUs or to compare OTUs
across studies.
To determine how well OptiFit performs for fitting new sequences to existing
OTUs, we employed a split dataset strategy, where each dataset was randomly
split into a reference fraction and a query fraction.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were fit to the _de novo_ OTUs with OptiFit.

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split).
OTU quality was highly similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to fit `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences fit to the Greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and 
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

@@

We developed a new algorithm for fitting sequences to existing OTUs and have
demonstrated its suitability for reference-based clustering.
OptiFit makes the iterative method employed by OptiClust available for tasks
where reference-based clustering is required.
We have shown that OTU quality is similar between OptiClust and OptiFit in open
reference mode, regardless of strategy employed.
Open reference OptiFit performs slower than OptiClust due to the additional _de
novo_ clustering step, so users may prefer OptiClust for tasks that do not
require reference OTUs.

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference databases, which are unlikely to
contain sequences from novel microbes.
This drop in quality was most notable with the RDP reference, which contained only 
`r ref_seqs[['rdp']]` sequences compared to `r ref_seqs[['silva']]` sequences in
SILVA and `r ref_seqs[['gg']]` in Greengenes.
Note that Greengenes has not been updated since 2013 at the time of this
writing, while SILVA and the RDP are updated regularly.
We recommend that users who require an independent reference database opt for
large databases with regular updates and good coverage of microbial diversity
for their environment.
Since OptiClust still performs faster than open reference OptiFit and creates
higher quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever consistent
OTUs are not required.

The OptiClust and OptiFit algorithms produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to map more sequences to OTUs than OptiFit in closed
reference mode.
While both OptiFit and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
OptiFit.
The OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single centroid sequence that seeded the OTU.
Additionally, OTU assignments clustered by VSEARCH are dependent on the order of
the input sequences, because each query is assigned to the OTU of the first
centroid sequence that is found within the distance threshold. TODO: is this true of VSEARCH or only USEARCH?
Because of this, VSEARCH sacrifices OTU quality by allowing more sequences to
fit to OTUs.

When fitting with the split dataset strategy, OTU quality was remarkably similar
when reference sequences were selected by a simple random sample or weighted by
abundance, but quality was slightly worse when sequences were weighted by
similarity. 
We recommend using a simple random sample since the more
sophisticated reference selection methods do not offer any benefit. 
The similarity in OTU quality between OptiClust and OptiFit with this strategy
demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
such as when comparing OTUs across studies.
However, when consistent OTUs are not required, we recommend using OptiClust for
_de novo_ clustering over the split strategy with OptiFit since OptiClust is
simpler to execute but performs similarly in terms of both run time and OTU
quality.

We have developed a new clustering algorithm that allows users to produce high
quality OTUs using already existing OTUs as a reference.
Unlike existing reference-based methods that map query sequences to a single
centroid sequence in each reference OTU, OptiFit considers all sequences in each
reference OTU when fitting query sequences, resulting in OTUs of a similar high
quality as those produced by the _de novo_ OptiClust algorithm.
Potential applications include fitting sequences to reference databases,
comparing taxonomic composition of microbiomes across different studies, 
or using OTU-based machine learning models to make predictions on new data.
OptiFit fills the missing option for fitting query sequences to existing OTUs
that does not sacrifice OTU quality for consistency of OTU assignments.

@@

We downloaded 16S rRNA gene amplicon sequences from four published datasets
isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
These datasets contain sequences from the V4 region of the 16S rRNA gene and
represent a selection of the broad types of natural communities that microbial
ecologists study.
We processed the raw sequences using mothur according to the Schloss Lab MiSeq
SOP and accompanying study by Kozich _et al._
[@schloss_miseq_nodate; @kozich_development_2013].
These steps included trimming and filtering for quality, aligning to the SILVA
reference alignment [@quast_silva_2013], discarding sequences that aligned
outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
and calculating distances between all pairs of sequences within each dataset
prior to clustering.

@@

To generate reference OTUs from independent databases, we downloaded sequences
from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
Project (v16) [@cole_ribosomal_2014].
These sequences were processed using the same steps outlined above followed by
clustering sequences into _de novo_ OTUs with OptiClust.
Processed reads from each of the four datasets were clustered with OptiFit to
the reference OTUs generated from each of the three databases.
When reference clustering with VSEARCH, processed datasets were fit directly to
the unprocessed Greengenes 97% OTU reference alignment, since this method is how
VSEARCH is typically used by the QIIME2 software reference-based clustering
[@bolyen_reproducible_2019; @noauthor_clustering_nodate].

>>>>

@@

Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
important step in characterizing the composition of microbial communities across
large datasets.
OptiClust, a _de novo_ OTU clustering method, has been
shown to produce higher quality OTU assignments than other methods and at
comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent
reference clustering methods is that OTU assignments from _de novo_ methods may
change when new sequences are added to a dataset.
However, in some cases one may wish to incorporate new samples into a previously
clustered dataset without performing clustering again on all sequences, such as
when comparing across datasets or deploying machine learning models where OTUs
are features.
Existing reference-based clustering methods produce consistent OTUs, but they
only consider the similarity of each query sequence to a single reference
sequence in an OTU, thus resulting in some dissimilar sequences being clustered
to the same OTU.
To provide an efficient and robust method to fit amplicon sequence data to
existing OTUs, we developed the OptiFit algorithm.
Like OptiFit, OptiClust considers the similarity of all pairs of reference and
query sequences in an OTU to produce OTUs of the best possible quality.
We tested OptiFit using four microbiome datasets with two different strategies:
by clustering to an external reference database or by splitting the dataset into a 
reference and query set and clustering the query sequences to the reference set
after clustering it using OptiClust.
The result is an improved implementation of closed and open-reference clustering.
OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
using the split dataset strategy, although the OTU quality and processing speed
depends on the database chosen when using the external database strategy.
OptiFit provides a suitable option for users who require consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering algorithm which uses the distance score
between all pairs of sequences in the dataset to cluster them into OTUs by
maximizing the Matthews Correlation Coefficient (MCC)
[@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_
methods such as the greedy clustering algorithms implemented in USEARCH and
VSEARCH, which only consider the distance between the query sequence and a
representative centroid sequence in the OTU [@edgar_search_2010;
@rognes_vsearch_2016].
In methods employing greedy clustering algorithms, only the distance between
each sequence and the centroid sequence is considered while clustering.
As a result, pairs of sequences in the same OTU may have a greater distance than
the specified threshold, i.e. they are false positives.
In contrast, the OptiClust algorithm takes into account the distance between all
pairs of sequences when considering how to cluster sequences into OTUs and is
thus less willing to take on false positives.
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow for
very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
Query sequences are then clustered into OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
Just like in _de novo_ clustering, reference-based methods that use greedy
clustering algorithms only consider the distance between each query and a single
reference sequence, causing some sequences in the same OTU to be more dissimilar
to each other than the specified threshold.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton. 
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
The MCC uses all values from a confusion matrix and ranges from zero to one,
with a score of one occurring when all sequence pairs are true positives and
true negatives, and a score of zero when all pairs are false positives and false
negatives.
Sequence pairs that are similar to each other (i.e. within the distance
threshold) are counted as true positives if they are clustered into the same OTU,
and false negatives if they are not in the the same OTU.
Sequence pairs that are not similar to each other are true negatives if they are
not clustered into the same OTU, and false positives if they are not in the same
OTU.
OptiClust iterations continue until the MCC stabilizes or until a maximum number
of iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.

@@

OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to cluster to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03) 
(Figure \ref{fig:algorithm}).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than clustered into a reference OTU.
All query and reference sequence pairs are considered when calculating the MCC.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be clustered into
reference OTUs are discarded, and the results only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) cluster the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then cluster the query sequences to the reference OTUs.

@@

While _de novo_ clustering produces the highest quality OTUs, researchers may
prefer to perform reference clustering to a public database because
reference-based methods produce consistent OTUs and are generally faster than
_de novo_ methods.
In closed reference mode, sequences that cannot be clustered into reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we clustered each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% using the V4 region of each
sequence (see Figure \ref{fig:workflow}).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
respectively.
clustering sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
was `r closed_fit_rdp_mcc` when clustering to RDP 
(see Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

@@

Since closed reference clustering does not cluster query sequences that could
not be clustered into reference OTUs, an additional measure of clustering
performance to consider is the fraction of query sequences that were able to be
clustered.
On average, more sequences were clustered with Greengenes as the reference 
(`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or with the RDP 
(`r frac_fit_rdp`%).
This mirrored the result reported above that Greengenes produced better OTUs in
terms of MCC score than either SILVA or RDP.
Note that _de novo_ and open reference clustering methods always cluster 100% of
sequences into OTUs.
The database chosen affects the final OTU assignments considerably in terms of
both MCC score and fraction of query sequences that could be clustered into the
reference OTUs.

@@

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH in terms of MCC, but performed 
`r sec_opticlust_vs_vsearch`% slower than VSEARCH.
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH, but VSEARCH was able to cluster
`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
reference database.
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of sequences in an OTU. <!-- VSEARCH "diameter" vs OptiFit "radius"
-->
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively).
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average respectively.
Thus, the more stringent OTU definition employed by OptiFit, which prefers the
query sequence to be similar to all other sequences in the OTU rather than to only one
sequence, resulted in fewer sequences being clustered to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTU quality may be poor when the reference database consists of sequences that
are too unrelated to the samples of interest, such as when samples contain novel
populations.
While _de novo_ clustering overcomes the quality limitations of reference
clustering to databases, OTU assignments are not consistent when new sequences
are added.
Researchers may wish to cluster new sequences to existing OTUs or to compare OTUs
across studies.
To determine how well OptiFit performs for clustering new sequences to existing
OTUs, we employed a split dataset strategy, where each dataset was randomly
split into a reference fraction and a query fraction.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were clustered to the _de novo_ OTUs with OptiFit.

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split).
OTU quality was highly similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences clustered to the Greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and 
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

@@

We developed a new algorithm for clustering sequences to existing OTUs and have
demonstrated its suitability for reference-based clustering.
OptiFit makes the iterative method employed by OptiClust available for tasks
where reference-based clustering is required.
We have shown that OTU quality is similar between OptiClust and OptiFit in open
reference mode, regardless of strategy employed.
Open reference OptiFit performs slower than OptiClust due to the additional _de
novo_ clustering step, so users may prefer OptiClust for tasks that do not
require reference OTUs.

When clustering to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be clustered into OTUs across any
dataset/database combination. 
This may reflect limitations of reference databases, which are unlikely to
contain sequences from novel microbes.
This drop in quality was most notable with the RDP reference, which contained only 
`r ref_seqs[['rdp']]` sequences compared to `r ref_seqs[['silva']]` sequences in
SILVA and `r ref_seqs[['gg']]` in Greengenes.
Note that Greengenes has not been updated since 2013 at the time of this
writing, while SILVA and the RDP are updated regularly.
We recommend that users who require an independent reference database opt for
large databases with regular updates and good coverage of microbial diversity
for their environment.
Since OptiClust still performs faster than open reference OptiFit and creates
higher quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than clustering to a database whenever consistent
OTUs are not required.

The OptiClust and OptiFit algorithms produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to cluster more sequences to OTUs than OptiFit in closed
reference mode.
While both OptiFit and VSEARCH use a distance or similarity threshold for
determining how to cluster sequences into OTUs, VSEARCH is more permissive than
OptiFit.
The OptiFit and OptiClust algorithms use all of the sequences to define an OTU,
preferring that all pairs of sequences (including reference and query sequences)
in an OTU are within the distance threshold in order to maximize the MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single centroid sequence that seeded the OTU.
Because of this, VSEARCH sacrifices OTU quality by allowing more sequences to
be clustered into OTUs.

When clustering with the split dataset strategy, OTU quality was remarkably similar
when reference sequences were selected by a simple random sample or weighted by
abundance, but quality was slightly worse when sequences were weighted by
similarity. 
We recommend using a simple random sample since the more
sophisticated reference selection methods do not offer any benefit. 
The similarity in OTU quality between OptiClust and OptiFit with this strategy
demonstrates the suitability of using OptiFit to cluster sequences to existing OTUs,
such as when comparing OTUs across studies.
However, when consistent OTUs are not required, we recommend using OptiClust for
_de novo_ clustering over the split strategy with OptiFit since OptiClust is
simpler to execute but performs similarly in terms of both run time and OTU
quality.

We have developed a new clustering algorithm that allows users to produce high
quality OTUs using already existing OTUs as a reference.
Unlike existing reference-based methods that cluster query sequences to a single
centroid sequence in each reference OTU, OptiFit considers all sequences in each
reference OTU when clustering query sequences, resulting in OTUs of a similar high
quality as those produced by the _de novo_ OptiClust algorithm.
Potential applications include clustering sequences to reference databases,
comparing taxonomic composition of microbiomes across different studies, 
or using OTU-based machine learning models to make predictions on new data.
OptiFit fills the missing option for clustering query sequences to existing OTUs
that does not sacrifice OTU quality for consistency of OTU assignments.

@@

We downloaded 16S rRNA gene amplicon sequences from four published datasets
isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
These datasets contain sequences from the V4 region of the 16S rRNA gene and
represent a selection of the broad types of natural communities that microbial
ecologists study.
We processed the raw sequences using mothur according to the Schloss Lab MiSeq
SOP [@schloss_miseq_nodate] and accompanying study by Kozich _et al._
[@kozich_development_2013].
These steps included trimming and filtering for quality, aligning to the SILVA
reference alignment [@quast_silva_2013], discarding sequences that aligned
outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
and calculating distances between all pairs of sequences within each dataset
prior to clustering.

@@

To generate reference OTUs from independent databases, we downloaded sequences
from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
Project (v16) [@cole_ribosomal_2014].
These sequences were processed using the same steps outlined above followed by
clustering sequences into _de novo_ OTUs with OptiClust.
Processed reads from each of the four datasets were clustered with OptiFit to
the reference OTUs generated from each of the three databases.
When reference clustering with VSEARCH, processed datasets were clustered directly to
the unprocessed Greengenes 97% OTU reference alignment, since this method is how
VSEARCH is typically used by the QIIME2 software reference-based clustering
[@bolyen_reproducible_2019; @noauthor_clustering_nodate].
